Gravitational Lens Modelling in a Citizen Science Context

### 2014-10-28 text/ms.tex

The first work on lens modelling
\citep{1981ApJ...244..723Y,1981ApJ...244..736Y} was developed after
the discovery of the first two gravitational lenses
\citep{1979Natur.279..381W,1980Natur.285..641W}, where a massive
galaxy causes a background quasar to appear as two or four
images.  For the second lens to be discovered (PG1115+080), mass
models scored an early success with the prediction that one of the
lensed images seen would split further into two at higher resolution.
That galaxies must sometimes cause multiple images had long been
expected \citep{1937ApJ....86..217Z}, and it had even been argued that
the phenomenon could help measure cosmological parameters
\citep{1964MNRAS.128..307R,1966MNRAS.132..101R}, but apparently nobody
was expecting that lenses would need detailed modelling.  The first
observations, however, immediately stimulated models.  The reason for
that lies in the image separation. Recall that image separations are
of order of the angular Einstein radius
\begin{equation}
\Theta_{\rm E}
\sim \left(\frac{4GM}{c^2 D_L}\right)^{1/2}
\simeq 0.1'' \left(\frac{M}{M_\odot}\right)^{1/2}
             \left(\frac{D_L}{\rm pc}\right)^{-1/2},
\end{equation}
where $D_L$ is the distance to the lens, and $M$ its mass.  A lensing
galaxy with $M\sim10^{11}M_\odot$ at $\sim1\rm\,Gpc$ would cause image
separations of $\sim1''$, which is comparable to the size of the
galaxy; typically the lensed images are seen through the galaxy halo.
Hence, the lensed images depend on the detailed mass distribution of
the lensing galaxy.  Galaxy lenses therefore require models of their
mass distributions.

>>>>

The first work on lens modelling
\citep{1981ApJ...244..723Y,1981ApJ...244..736Y} was developed after
the discovery of the first two gravitational lenses
\citep{1979Natur.279..381W,1980Natur.285..641W}, where a massive
galaxy causes a background quasar to appear as two or four
images.  For the second lens to be discovered (PG1115+080), mass
models scored an early success with the prediction that one of the
lensed images seen would split further into two at higher resolution.
That galaxies must sometimes cause multiple images had long been
expected \citep{1937ApJ....86..217Z}, and it had even been argued that
the phenomenon could help measure cosmological parameters
\citep{1964MNRAS.128..307R,1966MNRAS.132..101R}, but apparently nobody
was expecting that lenses would need detailed modelling.  The first
observations, however, immediately stimulated models.  The reason for
that lies in the image separation. Recall that image separations are
of order of the angular Einstein radius
\begin{equation}
\Theta_{\rm E}
\sim \left(\frac{4GM}{c^2 D_L}\right)^{1/2}
\simeq 0.1'' \left(\frac{M}{M_\odot}\right)^{1/2}
             \left(\frac{D_L}{\rm pc}\right)^{-1/2},
\end{equation}
where $D_L$ is the distance to the lens, and $M$ its mass.  A lensing
galaxy with $M\sim10^{11}\rm\,M_\odot$ at $\sim1\rm\,Gpc$ would cause image
separations of $\sim1''$, which is comparable to the size of the
galaxy; typically the lensed images are seen through the galaxy halo.
Hence, the lensed images depend on the detailed mass distribution of
the lensing galaxy.  Galaxy lenses therefore require models of their
mass distributions.
